---
title: "R_Practical_1_Answers"
author: "Lucy Sinke"
date: "20 October 2019"
output: 
  md_document:
    variant: gfm
---

***

## Part 1: RStudio and R Markdown

***

#### Answers to Question 1:

* `na.rm` indicates whether missing (NA) values should be stripped before the `mean` is calculated
* `sqrt` accepts numeric or complex vectors and arrays

***

## Part 2: Vectors

Define numeric variable `x`
```{r}
x <- 1 + 1
x
```

Load `tidyverse`
```{r message=F, warning=F}
library(tidyverse)
```

Define numeric vector `y`
```{r}
y <- c(5, 6, 7)
y
```

Define character vector `z`
```{r}
z <- c("a", "b", "c")
z
```

Define logical vector `w`
```{r}
w <- c(T, T, F)
w
```

***

#### Answers to Question 2:

```{r}
y[3]
```

```{r}
z[c(1,3)]
```

```{r}
y + 0:x
```

```{r}
rev(z)
```

***

## Part 3: Data Frames

Combine `y` and `z` into data frame `df`
```{r}
df <- data.frame(col1 = y, col2 = z, row.names = c("row1", "row2", "row3"))
df
```

Add `col3` to `df`
```{r}
df$col3 <- w
df
```

Add `row4` to `df` to make `df3`
```{r}
df2 <- data.frame(col1 = 8, col2 = "d", col3 = F, row.names = "row4")
df3 <- rbind(df, df2)
df3
```

#### Answers to Question 3:

Add a new column with the name `col4` to data frame `df` with the values `1, 1, 1`
```{r}
df$col4 <- c(1,1,1)
df
```

Make a new data frame `df4`, which is a subset of data frame `df` but only contains columns `col1` and `col2`, and rows `row2` and `row3`
```{r}
df4 <- df[c("row2", "row3"), c("col1", "col2")]
df4
```

Calculate the sum of `col1` of data frame `df3`
```{r}
sum(df3["col1"])
```

Change the row names of data frame `df4` to `rowX` and `rowY`
```{r}
rownames(df4) <- c("rowX", "rowY")
df4
```

***

## Part 4: Real Data

Read in the FHS data
```{r}
fhs <- read.csv(url("https://raw.githubusercontent.com/molepi/Molecular-Data-Science/master/RIntro_practical/data.csv"))
```

Remove observations with missing values
```{r}
fhs <- fhs[complete.cases(fhs), ]
sum(is.na(fhs))
```

Define formula `f`
```{r}
f <- y ~ x + z
f
```

#### Question 4:

Make a table of current smoking status
```{r}
xtabs(~ CURSMOKE, fhs)
```

Make a proportion table of BP medication
```{r}
prop.table(xtabs(~ BPMEDS, fhs))
```

Make a table showing the number of smokers for each education level
```{r}
xtabs(~ CURSMOKE + EDUC, fhs)
```

***

## Part 5: Data Manipulation

Create a subset of FHS containing BP variables
```{r}
fhsBP <- select(fhs, SYSBP, DIABP, BPMEDS)
head(fhsBP)
```

Create a subset of FHS containing all variables but MI
```{r}
fhsCOV <- select(fhs, -MI)
```

Create a subset of FHS containing BP variables using the `contains()` function
```{r}
fhsBP <- select(fhs, contains("BP"))
```

Show only observations where `BMI` is greater or equal to 40
```{r}
fhsBMI40 <- filter(fhs, BMI >= 40)
```

Show only observations for female smokers
```{r}
fhsSmokeF <- filter(fhs, SEX == 'Female', CURSMOKE == 'Yes')
```

***

#### Answers to Question 5:

What is the sex and smoking status of the youngest diabetic? 
**Female non-smoker**
```{r}
fhs %>%
  select(SEX, CURSMOKE, AGE, DIABETES) %>%
  filter(DIABETES == "Yes") %>%
  arrange(AGE) %>%
  head(n = 1)
```

What is the average systolic blood pressure of smokers who currently take BP medication? 
**168 mmHg**
```{r}
fhs %>%
  filter(CURSMOKE == "Yes", BPMEDS == "Yes") %>%
  summarise(AVG_SYSBP = mean(SYSBP))
```

For diabetics who experienced MI, what is their mean age, BMI, and glucose level? 
**55.4 years, 27.9 kg/m^2, and 165 mg/dL**
```{r}
fhs %>%
  filter(DIABETES == "Yes", MI == "Yes") %>%
  summarise(AVG_AGE = mean(AGE),
            AVG_BMI = mean(BMI),
            AVG_GLU = mean(GLUCOSE))
```

How many overweight smokers are in each education category? Do you notice a trend?
**Yes, there are less smokers in higher education groups**
```{r}
fhs %>%
  filter(BMI >= 30, CURSMOKE == "Yes") %>%
  group_by(EDUC) %>%
  summarise(N_EDUC = n())
```

Now look at the total number of people in each education category alongside the percentage who are overweight smokers. Is there a trend now?
**No, when looking at the percentage there doesn't appear to be a trend**
```{r}
fhs %>%
  mutate(OVRSM = ifelse(BMI >= 30 & CURSMOKE == "Yes", 1, 0)) %>%
  group_by(EDUC) %>%
  summarise(EDUC_N = n(),
            OVRSM_N = sum(OVRSM),
            PERC = OVRSM_N * 100 / EDUC_N)
```

***