---
title: "R_Practical_2_Answers"
author: "Lucy Sinke"
date: "20 October 2019"
output: 
  md_document:
    variant: gfm
---

***

## Part 1: Introduction to ggplot2

Load in data again, keeping only complete cases
```{r}
fhs <- read.csv(url("https://raw.githubusercontent.com/molepi/Molecular-Data-Science/master/RIntro_practical/data.csv"))
fhs <- fhs[complete.cases(fhs), ]
```

Load `tidyverse`
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

***

#### Answers to Question 1:

A bar plot of MI, coloured by smoking status
```{r}
fhs %>%
  ggplot() +
  geom_bar(aes(MI, fill=CURSMOKE)) +
  ggtitle("Bar chart of MI coloured by smoking status") +
  xlab("MI status") 
```

A histogram of total cholesterol in overweight individuals
```{r}
fhs %>%
  filter(BMI >= 30) %>%
  ggplot() +
  geom_histogram(aes(TOTCHOL), binwidth = 5, colour="cadetblue3") +
  ggtitle("Histogram of total cholesterol in overweight individuals") +
  xlab("Total cholesterol (mg/dL)") +
  geom_vline(aes(xintercept=240), colour='red', size=1.5)
```

Four density plots of age for each education level
```{r}
fhs %>%
  ggplot() +
  geom_density(aes(AGE, fill=EDUC), show.legend = F) +
  facet_wrap(~EDUC, labeller=label_both) +
  ggtitle("Density plots of age for each education level") +
  xlab("Age (years)")
```

Two bar plots of sex in current and non-smokers for those who experienced MI
```{r}
fhs %>%
  filter(MI == "Yes") %>%
  ggplot() +
  geom_bar(aes(SEX, fill=SEX)) + 
  facet_wrap(~ CURSMOKE) +
  ggtitle("Bar chart of sex in current and non-smokers", subtitle="for those who experienced MI") +
  xlab("Sex")
```

***

## Part 2: Comparative plots

***

#### Answers to Question 2:

A scatter plot of glucose levels against BMI, coloured by diabetes status
```{r}
fhs %>%
  ggplot() +
  geom_point(aes(GLUCOSE, BMI, colour=DIABETES)) +
  ggtitle("Scatter plot of glucose levels against BMI", subtitle="coloured by diabetes status") +
  xlab("Glucose level (mg/dL)") +
  ylab("BMI (kg/m^2)")
```

A horizontal box plot of systolic BP by sex for people over the age of 60

```{r}
fhs %>%
  filter(AGE >= 60) %>%
  ggplot() +
  geom_boxplot(aes(SEX, SYSBP, fill=SEX), show.legend=F) +
  coord_flip() +
  ggtitle("Horizontal box plot of systolic BP by sex", subtitle="for people over the age of 60")
```

A scatter plot of total cholesterol against glucose for categories of BMI

```{r}
fhs %>%
  ggplot() +
  geom_point(aes(TOTCHOL, GLUCOSE), colour='grey30') +
  facet_wrap(~cut(BMI, c(10,20,25,30,100))) +
  ggtitle("Scatter plot of total cholesterol against glucose", subtitle="for categories of BMI") +
  xlab("Total cholesterol (mg/dL)") +
  ylab("Glucose levels (mg/dL)")
  
```

***

## Part 3: Statistical Tests

Define table of smoking against MI
```{r}
SMOKE_MI <- xtabs(~CURSMOKE+MI, fhs)
SMOKE_MI
```

Chi-square test
```{r}
chisq <- chisq.test(SMOKE_MI)
chisq
```

This shows we have strong evidence (p=0.001) that smoking status and MI are associated.

Looking at the Pearson residuals, we see that people who had MI contribute most to this association.

```{r}
round(chisq$residuals, 3)
```

***

Pearson's correlation coefficient
```{r}
cor <- cor.test(~ SYSBP + DIABP, fhs) 
cor
```

There is very strong evidence (p<0.001) that systolic and diastolic blood pressure are strongly correlated (r2=0.786).

***

T-test for age difference between MI
```{r}
tt <- t.test(AGE ~ MI, fhs) 
tt
```

There is very strong (p<0.001) evidence that people who have had MI are older than those who didn't.

***

Logistic regression
```{r}
glmfit <- glm(MI ~ BMI + AGE + SEX, fhs, family = "binomial") 
summary(glmfit)
```

Odds ratios can also be accessed.

```{r}
exp(coef(glmfit))
```

***

#### Answers to Question 3:

Is age correlated with total cholesterol?
**Yes, there is strong evidence (p<0.001) that age is mildly correlated (r=0.252) with total cholesterol**
```{r}
cor <- cor.test(~ AGE + TOTCHOL, fhs) 
cor
```

Do diabetes have significantly different glucose levels when compared to non-diabetics?
**Yes, there is strong evidence (t=-10.99, p<0.001) that diabetics have higher glucose levels than non-diabetics**
```{r}
tt <- t.test(GLUCOSE ~ DIABETES, fhs) 
tt
```

Is MI associated with total cholesterol in individuals over the age of 50?
**Yes, there is strong evidence (p=0.003) that total cholesterol is associated with an increased risk of MI. Increasing total cholesterol by 1 unit confers a 0.3% increased odds of MI**
```{r}
fhs50 <- filter(fhs, AGE > 50)
glmfit <- glm(MI ~ TOTCHOL, fhs50, family = "binomial") 
summary(glmfit)
exp(coef(glmfit))
```

***

## Part 4: Loops

***

#### Answers to Question 4

Write a loop that returns the square of the first 3 elements in a vector
```{r}
x <- c(1,2,3,4)

for (i in 1:3) {
  out <- paste("The square of element", i, "of the vector is", x[i]^2)
  print(out)
}
```

Write a loop that prints each column name in the FHS dataset, alongside the number of characters in it.
```{r}
for (i in colnames(fhs)) {
  out <- paste(i, "-", nchar(i))
  print(out)
}
```

Write a loop that tells you at what integer the product of all previous positive integers is over 5 million

```{r}
x <- 1
product <- 1

while (product < 5000000) {
  x <- x + 1
  product <- product * x
}

print(x)
```

***

## Part 5: Functions

***

#### Answers to Question 5

Write a function that takes x, y, and z as arguments and returns their product
```{r}
product <- function(x, y, z) {
  p <- x * y * z
  return(p)
}

product(1, 2, 3)
```

Write a function that returns the correlation of all variables in the FHS data with BMI

```{r}
correlation <- function(x) {
  results <- vector()
  
  for (i in colnames(x)) {
    cor <- cor.test(~ as.numeric(get(i)) + BMI, x)
    results <- c(results, cor$estimate)
    }
  results <- data.frame(corr = results, row.names = colnames(x))
  results
}

correlation(fhs)
```

A function that associates all variables in the FHS data with age
```{r}
linearAllAge <- function(x) {
  results <- data.frame()
  fhsCov <- select(x, -AGE)
  
  for (i in colnames(fhsCov)) {
    fit <- lm(AGE ~ get(i), x)
    results <- rbind(results, summary(fit)$coefficients[2, , drop=F])
  }
  
  results <- cbind(Covariate = colnames(fhsCov), results)
  
  results %>%
    arrange(`Pr(>|t|)`) %>%
    print
}

linearAllAge(fhs)
```

